Rates and delay times of type Ia supernovae in the Dark Energy Survey

### 2021-03-08 draft.tex

The progenitors of a number of core-collapse SNe (CCSNe) have been identified in high-resolution pre-explosion images \citep{Smartt2009,Eldridge2013}, allowing detailed analysis of the latter stages of massive star evolution. However, this approach has not yet been successful in identifying a progenitor of a SN Ia (although see \citet{McCully2014} for evidence of a progenitor for a type Iax SN). Conversely, it has been possible to search for surviving remnants of the binary system -- main sequence stars or white dwarves left behind at the SN location or kicked \citep{Shen2018} -- displaying tentative evidence for a DD scenario. Without an observation of a SN Ia progenitor system or remnant companion, there is as yet no direct evidence ruling out any progenitor channel from being a significant contribution to the total population of SNe Ia.

Another method of inferring progenitor channels is the study of the SN Ia delay time distribution (DTD). The DTD describes the rate at which SNe Ia occur as a function of the delay time $\tau$ since an episode of star-formation, and thus carries characteristic signatures of the progenitor channel (see \citealt{Wang2012} for a review). While the SD scenario is able to produce a broad range of functional forms of the DTD, most variants of the DD scenario predict a power-law of the form $\tau^{\beta}$, with $\beta \sim -1$ \citep[e.g][]{Ruiter2009,Mennekens2010}.

A power law DTD implies that most SNe Ia explode soon after star formation, but also that there is a non-neglible fraction that explode on the order of Gyrs after the star formation episode. The DTD is therefore non-trivial to measure and various techniques have been developed to infer it indirectly. One such approach is to measure global properties of SN Ia host galaxies. Simple examples of such analyses are comparisons of the rates of SNe Ia in hosts that have been split by some observational property, such as morphology or colour, and have shown that SN Ia rate per unit stellar mass (SNuM) is significantly larger in late-type spiral galaxies as well as in bluer galaxies \citep[e.g.][]{Mannucci2005}. These observations heavily imply that SNe Ia rates are dependent on recent or ongoing star-formation, and thus that the majority of SNe Ia explode after short delay times. Conversely, the SNuM in E/S0 and red galaxies is non-negligible, suggesting that there is a secondary, much older component to the DTD \citep{Sullivan2006,Li2011a,Smith2012}. The DTD is thus approximated by a two-component ("$A + B$") model, where $A$ and $B$ are the normalisations of the DTD for "prompt" (i.e. proportional to instantaneous SFR) and "tardy" (i.e. proportional to overall stellar mass) SNe, respectively.

A complementary technique involves measuring the evolution of the volumetric rate of SNe Ia as a function of redshift and comparing this evolution to the average cosmic star-formation history (CSFH) of the Universe \citep{Gal-Yam2004,Strolger2004,Dahlen2004,Dahlen2008, Graur2013, Frohmaier2019}. This technique (the volumetric rate method) is also applicable to galaxy clusters, for which it is assumed that the star-formation histories are strongly peaked at some past epoch and thus that the rate of SNe Ia in clusters as a function of redshift is a more direct measure of the DTD \citep{Maoz2004,Graur2014,Maoz2014}. These studies have, almost ubiquitously, found $\beta$ to be consistent with $-1$.

Instead of comparing \textit{volumetric} rates to the \textit{cosmic} SFH, it is also possible to estimate the SFH of individual galaxies through the modelling of their stellar populations via spectral energy distribution (SED) fitting. Works such as \citet{Maoz2011}, \citet{Maoz2012} and \citet{Graur2013} estimated the DTD by measuring SFHs for a sample of field galaxies and comparing them to the number of SNe detected in each galaxy (the SFHR method), and have led to results that suggest a DTD power law with $\beta$ consistent with $-1$ \citep[e.g.][]{Maoz2011,Graur2013}. \citet{Childress2014} showed that the $A+B$ parameterisation arises as a direct consequence of the combination of a power-law DTD with the average SFHs of galaxies. The recent advances of integral field spectroscopy (IFS) allows for an extension of the SFHR method, by reconstructing the SFH for hundreds or thousands of local regions of each SN Ia host galaxy. Using this method, \citet{Castrillo2020} find a power-law slope of $-1.1\pm0.3$, while \citet{Chen2021} find $-1.4\pm0.3$.

While the majority of observations based on populations point towards a DD scenario for most, if not all, SNe Ia, finding self-consistent explosion models that recreate the observed luminosity function as well as correlations between luminosity, light curve parameters, and host galaxy properties has proven difficult. In particular, simulations based around explosions of $M_{\mathrm{Ch}}$ WDs fail to reproduced the observed relationship between SN luminosity and light curve duration (the Phillips relation; \citealt{Phillips1993}) as well as low-luminosity "peculiar" objects known as SN 1991bg-like SNe \citep{Sim2013,Blondin2017}. In recent years, attention focussed on explosions of sub-$M_{\mathrm{Ch}}$ WDs triggered by double detonations has led to promising results \citep[e.g.][]{Chen2017,Chen2018,Townsley2019,Gronow2020}. An additional factor in support of the double-detonation model is that the SN luminosity is related to the mass of the primary WD which can be related to its age, thereby providing an explanation for observed relation between light curve stretch and stellar age \citep{Rigault2013,Rigault2018,Nicolas2020}.

In this work, we combine parts of the traditional methods and derive a new, independent measurement of the SN Ia DTD. Instead of measuring the volumetric rate of SNe Ia, we measure the rate per galaxy as a function of stellar mass, as per \citet{Sullivan2006,Smith2012}. We use the stellar mass assembly model of \citet{Childress2014} to predict the stellar age distribution of galaxies for a given stellar mass at the mean redshift of our SN sample, and then forward model the DTD by convolving it with the stellar age distribution and comparing the predicted rates to the observed rates.

We introduce the SN host and field galaxy samples in Section \ref{sec:data} and describe our handling of incompleteness in Section \ref{sec:incompleteness}. We present the SN Ia rate per galaxy in Section \ref{sec:rates}. We outline the modelling of star-formation histories and measurement of the DTD in Section \ref{sec:model}. In Section \ref{sec:split_x1_c} we investigate how the DTD differs among sub-populations of SNe with different light curve characteristics. We conclude in Section \ref{sec:conclusion}.
Where relevant, we assume a spatially flat $\Lambda$CDM cosmology with $\Omega_m = 0.3$ and $H_0 = 70 \mathrm{ km s}^{-1}\mathrm{Mpc}^{-1}$. Unless otherwise stated, we assume Gaussian measurement uncertainties quoted at the $1\sigma$ level, and we quote $1\sigma$ credible intervals on derived parameters. Magnitudes are quoted in the AB system \citep{Oke1983}.

>>>>

The progenitors of a number of core-collapse SNe (CCSNe) have been identified in high-resolution pre-explosion images \citep{Smartt2009,Eldridge2013}, allowing detailed analysis of the latter stages of massive star evolution. However, this approach has not yet been successful in identifying a progenitor of a SN Ia (e.g. \citealt{Graur2014a,Kelly2014,Graur2019}, although see \citet{McCully2014} for evidence of a progenitor for a type Iax SN). Conversely, it has been possible to search for surviving remnants of the binary system -- main sequence stars or WDs left behind at the SN location or kicked out at high velocity \citep[e.g.][]{Schaefer2012,RuizLapuente2012,Kerzendorf2018,Kerzendorf2019}. \citet{Shen2018} discovered three high-velocity WDs consistent with models of double-detonation SNe Ia. However, without an unambiguous observation of a SN Ia progenitor system or remnant companion, there is as yet no direct evidence ruling out any progenitor channel from providing a significant contribution to the total population of SNe Ia.

Another method of inferring progenitor channels is the study of the SN Ia delay time distribution (DTD). The DTD describes the rate at which SNe Ia occur as a function of the delay time $\tau$ since an episode of star-formation, and thus carries characteristic signatures of the progenitor channel (see \citealt{Wang2012} for a review of the theory, and \citet{Maoz2017} for a review of observations). The SD scenario produces a broad range of functional forms for the DTD, most of which fail to account for long delay times \citep{Graur2014}. On the other hand, most variants of the DD scenario predict a power-law of the form $\tau^{\beta}$, with $\beta \sim -1$ \citep[e.g][]{Ruiter2009,Mennekens2010}.

As a statistical distribution acting on timescales from tens of Myr to several Gyr, the DTD is non-trivial to measure and various techniques have been developed to infer it indirectly. One such approach is to measure global properties of SN Ia host galaxies. Simple examples of such analyses are comparisons of the rates of SNe Ia in hosts that have been split by some observational property, such as morphology or colour, and have shown that SN Ia rate per unit stellar mass (SNuM) is significantly larger in late-type spiral galaxies as well as in bluer galaxies \citep[e.g.][]{Mannucci2005}. These observations were interpreted as showing that SNe Ia are strongly influenced by recent or ongoing star-formation, and thus that the majority of SNe Ia explode after short delay times. Conversely, the SNuM in E/S0 and red galaxies is non-negligible, suggesting that there is a secondary, much older component to the DTD \citep{Sullivan2006,Li2011a,Smith2012}. The DTD was thus approximated to first order by a two-component ("$A + B$") model, where $A$ and $B$ are the normalisations of the DTD for "prompt" (i.e. proportional to instantaneous SFR) and "tardy" (i.e. proportional to overall stellar mass) SNe, respectively.

A complementary technique involves measuring the evolution of the volumetric rate of SNe Ia as a function of redshift and comparing this evolution to the average cosmic star-formation history (CSFH) of the Universe \citep{Gal-Yam2004,Strolger2004,Dahlen2004,Dahlen2008, Graur2011,Graur2013, Rodney2014,Frohmaier2019}. This technique (the volumetric rate method) is also applicable to galaxy clusters, for which it is assumed that the star-formation histories are strongly peaked at some past epoch and thus that the rate of SNe Ia in clusters as a function of redshift is a more direct measure of the DTD \citep{Maoz2004,Maoz2010,Friedman2018,Freundlich2020}. These studies have, almost ubiquitously, found $\beta$ to be consistent with $-1$.

Instead of comparing \textit{volumetric} rates to the \textit{cosmic} SFH, it is also possible to estimate the SFH of individual galaxies through the modelling of their stellar populations via spectral energy distribution (SED) fitting. Works such as \citet{Maoz2011}, \citet{Maoz2012}, \citet{Graur2013}, and \citet{Graur2015} estimated the DTD by measuring SFHs for a sample of field galaxies and comparing them to the number of SNe detected in each galaxy (the SFHR method), and have led to results that suggest a DTD power law with $\beta$ consistent with $-1$.

\citet{Childress2014} showed that the $A+B$ parameterisation arises as a direct consequence of the combination of a power-law DTD with the average SFHs of galaxies. The recent advances of integral field spectroscopy (IFS) allows for an extension of the SFHR method, by reconstructing the SFH for hundreds or thousands of local regions of each SN Ia host galaxy. Using this method, \citet{Castrillo2020} find a power-law slope of $-1.1\pm0.3$, while \citet{Chen2021} find $-1.4\pm0.3$.

The majority of observational evidence based on studies of SN Ia populations thus points towards a DD scenario for most, if not all, SNe Ia. However, finding self-consistent explosion models that recreate the observed luminosity function as well as correlations between luminosity, light curve parameters, and host galaxy properties has proven difficult. In particular, simulations based around explosions of $M_{\mathrm{Ch}}$ WDs fail to reproduce the observed relationship between SN luminosity and light curve duration as well as "peculiar" objects \citep{Sim2013,Blondin2017}. In recent years, attention directed towards explosions of sub-$M_{\mathrm{Ch}}$ WDs triggered by double detonations has led to promising results \citep[e.g.][]{Shen2017,Shen2018,Townsley2019,Gronow2020,Shen2021} although they still struggle to match observations at late times in the light curve evolution \citep{Shen2021,Gronow2021}. An additional factor in support of the double-detonation model is that the SN luminosity is related to the mass of the primary WD which can be related to its age, thereby providing an explanation for observed relation between light curve stretch and stellar age \citep{Rigault2013,Rigault2018,Nicolas2020}.

In this work, we combine parts of the traditional methods and derive a new measurement of the SN Ia DTD. Instead of measuring the volumetric rate of SNe Ia, we measure the rate per galaxy as a function of stellar mass, as per \citet{Sullivan2006,Smith2012}. We use the stellar mass assembly model of \citet{Childress2014} to predict the stellar age distribution of galaxies for a given stellar mass at the mean redshift of our SN sample, and then forward model the DTD by convolving it with the stellar age distribution and comparing the predicted rates to the observed rates.

In Section \ref{sec:data} we introduce our large sample of SNe Ia from the Dark Energy Survey as well as our deep sample of field galaxies. We describe our detailed handling of incompleteness in Section \ref{sec:incompleteness}. We present the SN Ia rate per galaxy in Section \ref{sec:rates}. We outline our modelling of star-formation histories and our novel measurement of the power-law DTD in Section \ref{sec:model}. In Section \ref{sec:split_x1_c} we investigate how the DTD differs among sub-populations of SNe with different light curve characteristics, in particular the stretch, and show stretch to be strongly dependent on progenitor age. We conclude in Section \ref{sec:conclusion} by discussing the implications of the result.
Where relevant, we assume a spatially flat $\Lambda$CDM cosmology with $\Omega_m = 0.3$ and $H_0 = 70 \mathrm{ km s}^{-1}\mathrm{Mpc}^{-1}$. Unless otherwise stated, we assume Gaussian measurement uncertainties quoted at the $1\sigma$ level, and we quote $1\sigma$ credible intervals on derived parameters. Magnitudes are quoted in the AB system \citep{Oke1983}.
